preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2020-09-28 Manuscript/manuscript.Rmd
Small edits
added numbers to additional files in body of manuscript

Hi-C, TAD, chromatin, machine learning, random forest, spatial clustering

@@

We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundaries of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We used Arrowhead-called TAD boundaries [@durand2016juicer] and published Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Additional file: Arrowhead Script, Additional file: Table S1<!--Supplementary Table 1-->). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell line-specific genome annotation data.

@@

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Additional file: Table S2<!--Supplementary Table 3-->). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

@@

A total of 77 cell line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS) (Additional file: Table S3<!--Supplementary Table 2-->). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Additional file: Figure S1<!--Supplementary Figure 1-->).

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy compared to the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Additional file: Figure S2A<!--Supplementary Figure 3A-->). Random forest models built on Peakachu-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Additional file: Figure S2B, S2C<!--Supplementary Figure 3B,3C-->). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Additional file: Figure S3A<!--Supplementary Figures 4A-->). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Additional file: Figure S3B<!--Supplementary Figures 4B-->). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each chromosome set. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Additional file: Figure S3C<!--Supplementary Figures 4C-->). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from GM12878 cell line resulted in an average AUC=0.792 (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to 0.795. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=0.881 and 0.874, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file: Figure S3D<!--Supplementary Figure 4D-->). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base-level resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with a genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based spatial clustering (DBSCAN) and partitioning around medoids (PAM) to predict the location of boundary regions and summit points (highest boundary probability) at base-level resolution (see Methods; Additional file: Figure S4<!--Supplementary Figure 2-->).

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Additional file: Table S4<!--Supplementary Table 4-->). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Additional file: Table S2<!--Supplementary Table 3-->). This can be attributed to Arrowhead's inflation of called TADs, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Additional file: Table S2<!--Supplementary Table 3-->). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

@@

Next, we evaluated the biological significance of *preciseTAD* boundary points. Signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (Figure 5C, 5D; Additional file: Figure S5A, S5B<!--Supplementary Figure 5A, 5B-->). Surprised by the poor signal distribution around Arrowhead boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead boundaries, in contrast to the relatively well-performing Peakachu- and SpectralTAD-called boundaries (Additional file: Figure S5C, S5D<!--Supplementary Figure 5C, D-->). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Additional file: Figure S6-S7<!--Supplementary Figures 6-7-->). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBSs (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Additional file: Figure S8<!--Supplementary Figure 8-->). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model (Additional file: Figure S9<!--Supplementary Figure 9-->). <!--We found that 88.8% and 95.8% [??? These numbers are different from what is shown on Supp Fig 9; !!!the %'s shown are (left region)/(left + middle); should I take out the percentages in the figure? ??? Still unclear, calculations don't make sense. TBD] of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). [??? After clarification, MD to rewrite the first two sentences, unclear] -->This is expected as loop boundaries detected by Peakachu are more abundant, while comparatively wide TAD boundaries detected by Arrowhead likely represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu ground truth data in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (Additional file: Figure S10A, S10B<!--Supplementary Figure 10A, 10B-->). Likewise, when using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlaps (Additional file: Figure S10C, S10D<!--Supplementary 10C, 10D-->). Furthermore, boundaries predicted on unseen annotation data exhibited similar level of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line (Additional file: Figure S10<!--Supplementary Figure 10-->). These results indicate that *preciseTAD* pre-trained models can be successfully used to predict domain boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Additional file: Table S5<!--Supplementary Table 5-->). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be susceptible to different values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us from investigating its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons, we opted to report balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data resampling.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 3 different thresholds ($t=(0.975,0.99,1.0)$) and 6 different $\epsilon$ values ($\epsilon=(1000,5000,10000,15000,20000,25000)$)). We found that the normalized overlaps - calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries - between top TFBS sites and flanked *preciseTAD* boundaries converged for combinations of $t=1.0$ and $\epsilon=10000$ (Additional file: Figure S11<!--Supplementary Figure 11-->).

@@

TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools were used for training. The autosomal genomic coordinates in the GRCh37/hg19 human genome assembly were considered. Arrowhead-defined TAD boundaries were called from the GM12878 and K562 cell line Hi-C data (MAPQ>0, 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions) using the default parameters (Additional file: Arrowhead Script). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Additional file: Table S1<!--Supplementary Table 1-->). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin states (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Additional file: Table 3<!--Supplementary Table 2-->).

@@

Cell line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the average signal strength of the corresponding annotation (*Peak Signal Strength*), the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin states lack signal values (Additional file: Figure S1<!--Supplementary Figure 1-->).

@@

To investigate whether we could alleviate the resolution limitations of conventional domain calling tools, we developed *preciseTAD*. This algorithm leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution (Additional file: Figure S4<!--Supplementary Figure 2-->).

@@

where $e_{ks} = \textbf{I}\{ r_{s} \in (b_{k}-f, b_{k}+f)\}$ is the number of elemental regions $r$ of predictor $p$ that overlap with each flanked boundary. We evaluated $NE$ for combinations of $t=\{0.975, 0.99, 1.0 \}$ and $\epsilon = \{1000, 5000, 10000, 15000, 20000, 25000 \}$. The heuristic of $\epsilon$ is that density-reachable bases with genomic distances less than $\epsilon$ should occupy the same designated cluster. The default combination was set to $t=1.0$ and $\epsilon=10000$ based on our tests (Additional file: Figure S11<!--Supplementary Figure 11-->).

>>>>

Hi-C, TAD, machine learning, random forest, DBSCAN

@@

We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundaries of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We used Arrowhead-called TAD boundaries [@durand2016juicer] and published Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Additional file 1: Arrowhead Script, Additional file 2: Table S1<!--Supplementary Table 1-->). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell line-specific genome annotation data.

@@

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Additional file 3: Table S2<!--Supplementary Table 3-->). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

@@

A total of 77 cell line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS) (Additional file 4: Table S3<!--Supplementary Table 2-->). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Additional file 5: Figure S1<!--Supplementary Figure 1-->).

@@

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy compared to the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Additional file 6: Figure S2A<!--Supplementary Figure 3A-->). Random forest models built on Peakachu-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Additional file 6: Figure S2B, S2C<!--Supplementary Figure 3B,3C-->). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin state data, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Additional file 7: Figure S3A<!--Supplementary Figures 4A-->). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Additional file 7: Figure S3B<!--Supplementary Figures 4B-->). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each chromosome set. To determine transcription factors most important for boundary prediction across all chromosomes, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found four transcription factors, CTCF, RAD21, SMC3, and ZNF143, being consistently predictive of TAD and loop boundaries (Figure 4C; Additional file 7: Figure S3C<!--Supplementary Figures 4C-->). We optimized our model by only considering these top four TFBS when building the random forest model, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from GM12878 cell line resulted in an average AUC=0.792 (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to 0.795. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=0.881 and 0.874, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D<!--Supplementary Figure 4D-->). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base-level resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with a genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based spatial clustering (DBSCAN) and partitioning around medoids (PAM) to predict the location of boundary regions and summit points (highest boundary probability) at base-level resolution (see Methods; Additional file 8: Figure S4<!--Supplementary Figure 2-->).

When trained using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb, respectively, the *preciseTAD* model predicted a total of 12,258 TAD and 15,707 chromatin loop boundaries in GM12878, as well as 9,603 TAD and 11,154 chromatin loop boundaries in K562 cell line (Additional file 9: Table S4<!--Supplementary Table 4-->). We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Additional file 3: Table S2<!--Supplementary Table 3-->). This can be attributed to Arrowhead's inflation of called TADs, that, when visualized, often do not correspond to domains enriched in internal interactions (Figure 5A) and signal of known drivers of domain boundaries (Figure 5B). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Additional file 3: Table S2<!--Supplementary Table 3-->). This can be attributed to Peakachu's use of only CTCF sites to call boundaries, while *preciseTAD* leverages four known drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant than those detected from Hi-C data alone.

@@

Next, we evaluated the biological significance of *preciseTAD* boundary points. Signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (Figure 5C, 5D; Additional file 10: Figure S5A, S5B<!--Supplementary Figure 5A, 5B-->). Surprised by the poor signal distribution around Arrowhead boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead boundaries, in contrast to the relatively well-performing Peakachu- and SpectralTAD-called boundaries (Additional file 10: Figure S5C, S5D<!--Supplementary Figure 5C, D-->). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone (Additional file 11-12: Figure S6-S7<!--Supplementary Figures 6-7-->). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBSs (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Additional file 13: Figure S8<!--Supplementary Figure 8-->). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model (Additional file 14: Figure S9<!--Supplementary Figure 9-->). <!--We found that 88.8% and 95.8% [??? These numbers are different from what is shown on Supp Fig 9; !!!the %'s shown are (left region)/(left + middle); should I take out the percentages in the figure? ??? Still unclear, calculations don't make sense. TBD] of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 9). [??? After clarification, MD to rewrite the first two sentences, unclear] -->This is expected as loop boundaries detected by Peakachu are more abundant, while comparatively wide TAD boundaries detected by Arrowhead likely represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu ground truth data in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (Additional file 15: Figure S10A, S10B<!--Supplementary Figure 10A, 10B-->). Likewise, when using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlaps (Additional file 15: Figure S10C, S10D<!--Supplementary 10C, 10D-->). Furthermore, boundaries predicted on unseen annotation data exhibited similar level of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line (Additional file 15: Figure S10<!--Supplementary Figure 10-->). These results indicate that *preciseTAD* pre-trained models can be successfully used to predict domain boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Additional file 16: Table S5<!--Supplementary Table 5-->). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be susceptible to different values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us from investigating its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons, we opted to report balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data resampling.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 3 different thresholds ($t=(0.975,0.99,1.0)$) and 6 different $\epsilon$ values ($\epsilon=(1000,5000,10000,15000,20000,25000)$)). We found that the normalized overlaps - calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries - between top TFBS sites and flanked *preciseTAD* boundaries converged for combinations of $t=1.0$ and $\epsilon=10000$ (Additional file 17: Figure S11<!--Supplementary Figure 11-->).

@@

TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools were used for training. The autosomal genomic coordinates in the GRCh37/hg19 human genome assembly were considered. Arrowhead-defined TAD boundaries were called from the GM12878 and K562 cell line Hi-C data (MAPQ>0, 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions) using the default parameters (Additional file 1: Arrowhead Script). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Additional file 2: Table S1<!--Supplementary Table 1-->). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin states (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Additional file 4: Table S3<!--Supplementary Table 2-->).

@@

Cell line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the average signal strength of the corresponding annotation (*Peak Signal Strength*), the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin states lack signal values (Additional file 5: Figure S1<!--Supplementary Figure 1-->).

@@

To investigate whether we could alleviate the resolution limitations of conventional domain calling tools, we developed *preciseTAD*. This algorithm leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution (Additional file 8: Figure S4<!--Supplementary Figure 2-->).

@@

where $e_{ks} = \textbf{I}\{ r_{s} \in (b_{k}-f, b_{k}+f)\}$ is the number of elemental regions $r$ of predictor $p$ that overlap with each flanked boundary. We evaluated $NE$ for combinations of $t=\{0.975, 0.99, 1.0 \}$ and $\epsilon = \{1000, 5000, 10000, 15000, 20000, 25000 \}$. The heuristic of $\epsilon$ is that density-reachable bases with genomic distances less than $\epsilon$ should occupy the same designated cluster. The default combination was set to $t=1.0$ and $\epsilon=10000$ based on our tests (Additional file 17: Figure S11<!--Supplementary Figure 11-->).
